Inferring the gravitational potential of the Milky Way with few, precisely measured stars

### 2014-01-26 text/notes/likelihood/likelihood.tex

Starting with notation, take ${\bf D}^{(i)}$ to be the observed, heliocentric 6D position of the $i$th star -- e.g., the measured position on the sky, ($l$, $b$), distance, $D$, proper motions, ($\mu_l$, $\mu_b$), and radial velocity, $v_r$ -- and ${\bf W}^{(i)}$ to be the corresponding \emph{true}, 6D position of the same star in the same coordinate and reference frame. We assume that this star was once part of a progenitor system -- e.g., satellite galaxy -- tidally disrupting and forming a cold debris structure in the potential, $\Phi$, of some parent galaxy. The satellite is observed to be at heliocentric 6D position ${\bf D}_s$, with true position ${\bf W}_s$, and shape described by a covariance matrix, $\V_s$ (the subscript $s$ refers to the satellite). The true, 6D spherical, heliocentric positions of the stars and satellite -- $\W^{(i)}$ and $\W_s$ -- have corresponding 6D cartesian, Galactocentric coordinate vectors $\X^{(i)}$ and $\X_s$.

We model the observed, present-day position of a star as the result of sampling from a time-dependent 6D distribution -- the progenitor system -- at some time $\tau^{(i)}$ in the past, evolved to present day in the parent potential, neglecting the effect of self-interaction with the halo of the progenitor. We model the progenitor as a 6D Gaussian with mean vector given by the position of the satellite at any given time, ${\bf W}_s(t)$, and with shape defined by the time-dependent covariance matrix $\V_s(t)$. With these assumptions we can write the joint posterior probability for the observed position of a star, properties of the disrupting satellite, and parent potential given the data as:
\begin{align}
	%p(\Phi, \W^{(i)}, \W_s, \V_s, \tau^{(i)} \given \D^{(i)}, \D_s) &= 
	%	\frac{1}{\mathcal{Z}} p(\D^{(i)} \given \W^{(i)}) p(\D_s \given \W_s) p(\W^{(i)} \given \W_s, \V_s, \tau^{(i)}, \Phi)
	p(\Phi, \W^{(i)}, \W_s, \V_s, \tau^{(i)} \given \D^{(i)}, \D_s) &= 
		\frac{1}{\mathcal{Z}} p(\D^{(i)} \given \W^{(i)}) p(\D_s \given \W_s) p(\D^{(i)} \given \D_s, \V_s, \tau^{(i)}, \Phi)
\end{align}
where the factor $\mathcal{Z}$ only depends on the properties of the data (the \emph{evidence}), and is thus constant when varying the model parameters. We assume that the observational errors are Gaussian:
\begin{align}
	p(\D^{(i)} \given \W^{(i)}) &= \mathcal{N}(\W^{(i)} \given \D^{(i)}, \bSigma^{(i)})\\
	p(\D_s \given \W_s) &= \mathcal{N}(\W_s \given \D_s, \bSigma_s)
\end{align}
where the covariance matrices $\bSigma^{(i)}$ and $\bSigma_s$ represent the observational variances on the observed 6D position of star and the satellite, respectively. [...] hmm...
\begin{align}
	p(\D^{(i)} \given \D_s, \V_s, \tau^{(i)}, \Phi) &= p(\W^{(i)} \given \W_s, \V_s, \tau^{(i)}, \Phi) \times \det{\bf J}^{(i)}\\
	&= \left(\mathcal{N}(\W^{(i)} \given \W_s, \V_s) \times \det{\bf J}^{(i)}\right)\rvert_{t=\tau^{(i)}}
\end{align}

>>>>

Take ${\bf D}$ to be the observed, heliocentric 6D position of a star --- e.g., the measured position on the sky, ($l$, $b$); distance, $D$; proper motions, ($\mu_l$, $\mu_b$); and line-of-sight velocity, $v_r$ --- and take ${\bf W}$ to be the corresponding \emph{true}, 6D position of the same star in the same coordinate and reference frame. We have determined that this star was once part of a progenitor system (e.g., satellite galaxy) that is tidally disrupting and forming a cold debris structure in the potential, $\Phi$, of some parent galaxy. The position of the satellite is observed to be at heliocentric 6D position ${\bf D}_s$, with true position ${\bf W}_s$, where the subscript $s$ refers to the satellite. The true, 6D spherical, heliocentric positions of the star and satellite -- $\W$ and $\W_s$ -- have corresponding 6D cartesian, Galactocentric coordinate vectors $\X$ and $\X_s$.

We model the present-day position and velocity of the disrupted star as the result of sampling from isotropic, log-normal distributions centered on the tidal radius, $R_{\rm tide}$, and escape velocity, $V_{\rm esc}$ of the satellite, respectively, at some time $\tau$ in the past, evolved to present day in the parent potential, $\Phi$. In this model, we neglect the effect of self-interaction with the halo of the progenitor [TODO: maybe we need to justify this, add APW's figure from paper 1 ref. response?].

With these assumptions we can write the joint posterior probability for the observed position of a star, properties of the disrupting satellite, and parent potential given the data as:
\begin{align}
	%p(\Phi, \W^{(i)}, \W_s, \V_s, \tau^{(i)} \given \D^{(i)}, \D_s) &= 
	%	\frac{1}{\mathcal{Z}} p(\D^{(i)} \given \W^{(i)}) p(\D_s \given \W_s) p(\W^{(i)} \given \W_s, \V_s, \tau^{(i)}, \Phi)
	p(\Phi, \W^{(i)}, \W_s, \V_s, \tau^{(i)} \given \D^{(i)}, \D_s) &= 
		\frac{1}{\mathcal{Z}} p(\D^{(i)} \given \W^{(i)}) p(\D_s \given \W_s) p(\D^{(i)} \given \D_s, \V_s, \tau^{(i)}, \Phi)
\end{align}
where the factor $\mathcal{Z}$ only depends on the properties of the data (the \emph{evidence}), and is thus constant when varying the model parameters. We assume that the observational errors are Gaussian:
\begin{align}
	p(\D^{(i)} \given \W^{(i)}) &= \mathcal{N}(\W^{(i)} \given \D^{(i)}, \bSigma^{(i)})\\
	p(\D_s \given \W_s) &= \mathcal{N}(\W_s \given \D_s, \bSigma_s)
\end{align}
where the covariance matrices $\bSigma^{(i)}$ and $\bSigma_s$ represent the observational variances on the observed 6D position of star and the satellite, respectively. [...] hmm...
\begin{align}
	p(\D^{(i)} \given \D_s, \V_s, \tau^{(i)}, \Phi) &= p(\W^{(i)} \given \W_s, \V_s, \tau^{(i)}, \Phi) \times \det{\bf J}^{(i)}\\
	&= \left(\mathcal{N}(\W^{(i)} \given \W_s, \V_s) \times \det{\bf J}^{(i)}\right)\rvert_{t=\tau^{(i)}}
\end{align}

### 2014-01-26 text/paper2/ms.tex

\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\gaia}{\project{Gaia}}
\newcommand{\given}{\,|\,}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\transpose}[1]{{#1}^{\mathsf{T}}}
\newcommand{\inverse}[1]{{#1}^{-1}}

@@

\title{Inferring the gravitational potential of the Milky Way with measurements of \emph{just two stars}}
\author{Adrian M. Price-Whelan, David W. Hogg, Kathryn V. Johnston, others}

\begin{abstract}
Informative phase-space structures, and in particular those created by
disrupting stellar systems or revealed by chemical tags, may deliver
very precise measures of the gravitational potential in the Milky Way
Halo.  Here we show that even a single pair of stars---two stars that
are known (for some non-kinematic reason) to be likely to be
associated with one another at birth---could provide a significant
constraint on the potential.  The inference is based on a
probabilistic generative model---a likelihood function and priors over
nuisance parameters---that evolves a past putative origin disruption
event forward in time to the present day.  The time, six-volume, and
phase-space location of the origin event are nuisance parameters in
the model and marginalized away.  The method makes no assumption of
integrability, works with finite or even large observational
uncertainties, does not require all dimensions of phase space to be
observed, handles non-zero probability that the two stars are not in
fact associated, and generalizes naturally to larger numbers of stars
and multiple independent structures.  Applications to the GD-1 cold
stellar stream and current surveys for RR Lyrae stars in the Halo are
discussed.
\end{abstract}

\keywords{
  this ---
  that ---
  Milky Way
}

@@

[Actions, angles, integrable, not, etc.  Can integrate in any
  potential, even a time-varying one.  APW \& KVJ have used this to
  ``run the clock back''.  Here we only let ourselves \emph{generate
    the data}.  Why is an objective based on past collisions not the
  same as assuming that the stars are on the same orbit?]

@@

\section{Experiments}

@@

[basic experiment: $N=2$, $K=1$, high $P_{nk}$, all dimensions
  measured at best-present-day quality]

[\gaia\ quality]

[low $P_{nk}$, MW bg model]

[unknown $P_{nk}$]

[missing data]

[$N>2$]

[$K>1$]

[Can we do a real experiment with two real RRL stars?]

>>>>

\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\gaia}{\project{Gaia}~}
\newcommand{\spitzer}{\project{Spitzer}~}
\newcommand{\given}{\,|\,}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\transpose}[1]{{#1}^{\mathsf{T}}}
\newcommand{\inverse}[1]{{#1}^{-1}}

@@

\title{Inferring the gravitational potential of the Milky Way with few, precisely measured stars}
\author{Adrian M. Price-Whelan, DWH, KVJ, others?}

\begin{abstract}
%Informative phase-space structures, and in particular those created by
%disrupting stellar systems or revealed by chemical tags, may deliver
%very precise measures of the gravitational potential in the Milky Way
%Halo.  Here we show that even a single pair of stars---two stars that
%are known (for some non-kinematic reason) to be likely to be
%associated with one another at birth---could provide a significant
%constraint on the potential.  The inference is based on a
%probabilistic generative model---a likelihood function and priors over
%nuisance parameters---that evolves a past putative origin disruption
%event forward in time to the present day.  The time, six-volume, and
%phase-space location of the origin event are nuisance parameters in
%the model and marginalized away.  The method makes no assumption of
%integrability, works with finite or even large observational
%uncertainties, does not require all dimensions of phase space to be
%observed, handles non-zero probability that the two stars are not in
%fact associated, and generalizes naturally to larger numbers of stars
%and multiple independent structures.  Applications to the GD-1 cold
%stellar stream and current surveys for RR Lyrae stars in the Halo are
%discussed.
TODO...
\end{abstract}

\keywords{
  Galaxy: structure
  ---
  Galaxy: halo
  ---
  cosmology: dark matter
}

@@

[Actions, angles, integrable, not, etc.  Can integrate in any
  potential, even a time-varying one. We use this to
  ``run the clock back''.  Why is an objective based on past collisions not the
  same as assuming that the stars are on the same orbit?]

@@

\section{Tests on simulated data}

@@

\subsection{\gaia + \spitzer}

\subsection{\spitzer + current data}
